---
title: "IRT optimism analysis"
author: "George"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
params:
  quick: false  # TRUE = 2 short chains per model, for smoke-testing the pipeline
  kfold: true   # grouped K-fold CV (slow the first time; results are cached)
---

The descriptive Bayesian model in Powell (CogSci 2022) is

$$\text{logit}(\mu_{ij}) = \text{logit}(x_{ij}) + (\alpha_i + \beta_i g_{ij})\,E_{ij},\qquad E_{ij} = \text{logit}(b_j) - \text{logit}(\tilde b_{ij})$$

where $x$ is the personal prior (`_P1`), $\tilde b$ the participant's estimate of the average person's risk (`_B1`), $b$ the true base rate shown (`item_prob`), and the response is the personal posterior (`_S2`). $g_{ij}$ codes good news.

Because $\text{logit}(x_{ij})$ enters with a fixed coefficient of 1, this is a Beta GLMM with an offset and a person-level random slope on $E$: `post ~ 0 + E + offset(logit_pre) + (0 + E | subj)`. In IRT terms, $\alpha_i$ is a person trait ("evidence sensitivity"), $E_{ij}$ is a known, trial-varying stimulus intensity, and the paper's model has no item parameters. The ladder below adds them.

| Model | Adds | IRT analogue |
|---|---|---|
| M0 | Rational model, $\alpha = 1$ | — |
| M1 | Person $\alpha_i$ (paper's `dbayes`) | Person-only (Rasch without items) |
| M2 | Person $\beta_i$, **correlated** with $\alpha_i$ (paper's `op_model` had them independent) | 2-dimensional person model |
| M3 | Item random slopes on $\alpha$ and $\beta$ | Crossed person × item (1PL-like) |
| M3b | LOT-R and age predicting $\alpha$, $\beta$ | Latent regression / explanatory IRT |
| M4 | Item loadings $\lambda_j$ on the person $\alpha$ trait | 2PL |
| M5 | Separate "update at all?" and "how much?" parts, correlated across parts | Two-part / hurdle IRT |
| M6 | Free weights on $b$ and $\tilde b$ | Tests the rational-model constraint |

```{r setup, include=FALSE}
library(tidyverse)
library(brms)
library(tidybayes)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# /local/ is gitignored; fits are cached there and refit only when the model changes
fit_dir <- if (params$quick) "local/irt-fits/quick" else "local/irt-fits"
dir.create(fit_dir, recursive = TRUE, showWarnings = FALSE)

fit_brm <- function(name, ...) {
  brm(
    ...,
    chains = if (params$quick) 2 else 4,
    iter = if (params$quick) 500 else 2000,
    cores = 4,
    seed = 2022,
    save_pars = save_pars(all = TRUE),
    file = file.path(fit_dir, name),
    file_refit = "on_change"
  )
}
```

```{r process-data, eval=F}
df_all <- read_csv("data/life-events-large-2021-12-14.csv") %>%
  filter(DistributionChannel=="anonymous") %>%
  rename(Duration = `Duration (in seconds)`) %>%
  mutate(subj_id = 1:n())

df_passing <- df_all %>%
  filter(Progress==100, check_77==77) %>%
  select(subj_id, contains("_B"), contains("_P"), contains("_S"), contains("lotr"), age, gender, marital)

items <- read_tsv("be a victim of a car break-in	76
die before the age of 80	41
suffer heart failure	31
die before the age of 70	18
suffer a stroke	17
be physically assaulted by a stranger	26
be diagnosed with any kind of cancer	40
become divorced	45
have cataracts	65
be diagnosed with skin cancer	3
become a homeowner	82
live past 90 years old	11
maintain a healthy weight	58
have an achievement recognized by the press	3
have the financial security to retire comfortably at 65	22
live past 100 years old	1
be in a marriage that lasts at least 10 years	61
be earning more than $75,000/yr within 5 years	16
be earning more than $100,000/yr within 5 years	9
make it through all of next year without catching a cold	23",
col_names = c("item_text", "item_prob")) %>%
  mutate(item_num = as.character(1:20))

df <- df_passing %>%
  mutate_at(vars(contains("lotr_")), ~recode(.x,
                            "Strongly agree" = 4,
                            "Somewhat agree" = 3,
                            "Neutral" = 2,
                            "Somewhat disagree" = 1,
                            "Strongly disagree" = 0)) %>%
  mutate(
    # LOT-R scored items are 1, 3, 4, 7, 9, 10 (3, 7, 9 reversed); 2, 5, 6, 8 are fillers.
    # processed_data.Rdata was saved with the paper's scoring, which also added fillers 6 and 8;
    # the corrected score is recomputed in the load-data chunk below.
    lotr = lotr_1 + (4-lotr_3) + lotr_4 + (4-lotr_7) + (4-lotr_9) + lotr_10
    ) %>%
  select(-contains("lotr_")) %>%
  gather(trial, response, contains("_B"), contains("_P"), contains("_S")) %>%
  mutate(
    trial_type = case_when(
      grepl("_B1", trial) ~ "baserate",
      grepl("_P1", trial) ~ "pre",
      grepl("_S2", trial) ~ "post"
    )
  ) %>%
  mutate(item_num = str_match(trial, "([0-9]*)_")[,2]) %>%
  left_join(items, by = "item_num") %>%
    mutate(response = response/100, item_prob = item_prob/100) %>%
  mutate(response = case_when(
    response == 0 ~ .005, # maybe .001? .0025?
    response == 1 ~ .995,
    TRUE ~ response
  )) %>%
  select(-trial) %>%
  spread(trial_type, response) %>%
  group_by(subj_id) %>%
  mutate(
    logit_base = qlogis(baserate),
    logit_pre = qlogis(pre),
    logit_truebase = qlogis(item_prob),
    logit_post = qlogis(post),
    update = post - pre
  )

df_reg <- df %>%
  mutate(
    item_num = as.numeric(item_num),
    pos_item = if_else(item_num > 10, 1, 0),
    good_news = case_when(
      pos_item==1 & item_prob > baserate ~ 1,
      pos_item==0 & item_prob < baserate ~ 1,
      TRUE ~ 0
    )
  )

save(df, df_reg, df_all, df_passing, items,
     file="processed_data.Rdata")
```

# Data

Columns per item: `_B1` = estimate of the average person's probability ($\tilde b$), `_P1` = own prior ($x$), `_S2` = own posterior after seeing the true base rate. `lotr` is the Life Orientation Test–Revised (trait optimism).

```{r load-data}
load("processed_data.Rdata")

lotr_scores <- df_passing %>%
  select(subj_id, lotr_1, lotr_3, lotr_4, lotr_7, lotr_9, lotr_10) %>%
  mutate(across(-subj_id, ~ recode(.x,
                                   "Strongly agree" = 4,
                                   "Somewhat agree" = 3,
                                   "Neutral" = 2,
                                   "Somewhat disagree" = 1,
                                   "Strongly disagree" = 0))) %>%
  transmute(subj_id,
            lotr_fixed = lotr_1 + (4 - lotr_3) + lotr_4 + (4 - lotr_7) + (4 - lotr_9) + lotr_10)

d <- df_reg %>%
  ungroup() %>%
  left_join(lotr_scores, by = "subj_id") %>%
  mutate(
    E = logit_truebase - logit_base,  # log evidence ratio carried by the true base rate
    g = good_news,
    valence = if_else(pos_item == 1, "positive event", "negative event"),
    upd = as.integer(abs(post - pre) > 1e-9),  # changed their estimate at all
    upd_l = upd == 1,                          # logical copy for brms subset()
    item = as.character(item_num),
    subj = factor(subj_id),
    lotr_z = as.numeric(scale(lotr_fixed)),
    age_z = as.numeric(scale(age))
  )

item_labels <- items %>% transmute(item = as.character(as.numeric(item_num)), item_text, item_prob)

c(n_subj = n_distinct(d$subj), n_obs = nrow(d),
  prop_no_change = mean(d$upd == 0), prop_good_news = mean(d$g))
```

Good news is largely determined by the item, because perceived base rates are compressed toward 50%. So $\beta$ is partly identified from between-item contrasts unless items get their own parameters.

```{r good-news-by-item}
d %>%
  group_by(item) %>%
  summarize(p_good = mean(g), mean_E = mean(E), p_no_change = mean(upd == 0)) %>%
  left_join(item_labels, by = "item") %>%
  select(item, item_text, item_prob, p_good, mean_E, p_no_change) %>%
  knitr::kable(digits = 2)
```

# Model ladder

Priors follow the paper: $\alpha \sim N(1, .5)$, $\beta \sim N(0, .5)$. Group-level SDs get half-$N(0, .5)$, a weak constraint that matters mostly for the 20-item SDs.

## M0: rational model

```{r m0}
m0 <- fit_brm(
  "m0_rational",
  post ~ 0 + E + offset(logit_pre),
  data = d, family = Beta(),
  prior = set_prior("constant(1)", class = "b", coef = "E")
)
```

## M1: person evidence weight $\alpha_i$

```{r m1}
m1 <- fit_brm(
  "m1_person_alpha",
  post ~ 0 + E + offset(logit_pre) + (0 + E | subj),
  data = d, family = Beta(),
  prior = set_prior("normal(1, .5)", class = "b", coef = "E") +
    set_prior("normal(0, .5)", class = "sd")
)
```

```{r m1-summary}
summary(m1)
```

## M2: person $\alpha_i$ and $\beta_i$, correlated

```{r m2}
priors_ab <- set_prior("normal(1, .5)", class = "b", coef = "E") +
  set_prior("normal(0, .5)", class = "b", coef = "E:g") +
  set_prior("normal(0, .5)", class = "sd")

m2 <- fit_brm(
  "m2_person_alpha_beta",
  post ~ 0 + E + E:g + offset(logit_pre) + (0 + E + E:g | subj),
  data = d, family = Beta(),
  prior = priors_ab
)
```

```{r m2-summary}
summary(m2)
```

## M3: crossed person × item

```{r m3}
m3 <- fit_brm(
  "m3_crossed",
  control = list(adapt_delta = .95),
  post ~ 0 + E + E:g + offset(logit_pre) + (0 + E + E:g | subj) + (0 + E + E:g | item),
  data = d, family = Beta(),
  prior = priors_ab
)
```

```{r m3-summary}
summary(m3)
```

Item-level $\alpha_j$ (updating on bad news) and $\alpha_j + \beta_j$ (good news):

```{r m3-items, fig.width=8, fig.height=6}
m3_items <- m3 %>%
  spread_draws(b_E, `b_E:g`, r_item[item, term]) %>%
  pivot_wider(names_from = term, values_from = r_item) %>%
  mutate(`bad news` = b_E + E, `good news` = b_E + E + `b_E:g` + `E:g`) %>%
  pivot_longer(c(`bad news`, `good news`), names_to = "news", values_to = "alpha") %>%
  mutate(item = as.character(item)) %>%
  left_join(item_labels, by = "item") %>%
  left_join(distinct(d, item, valence), by = "item")

m3_items %>%
  ggplot(aes(y = reorder(item_text, alpha), x = alpha, color = news)) +
  stat_pointinterval(position = position_dodge(width = .5), .width = .95) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  facet_grid(valence ~ ., scales = "free_y", space = "free_y") +
  labs(x = "Evidence weight", y = NULL, color = NULL) +
  theme_bw() +
  theme(panel.grid = element_blank(), legend.position = "bottom")
```

## M3b: latent regression on LOT-R and age

This replaces the paper's two-stage correlation of point estimates (Fig. 4) with person covariates inside the model. Participants with missing age are dropped, so M3b is left out of the LOO comparison.

```{r m3b}
m3b <- fit_brm(
  "m3b_latent_regression",
  control = list(adapt_delta = .95),
  post ~ 0 + E + E:g + E:lotr_z + E:g:lotr_z + E:age_z + E:g:age_z + offset(logit_pre) +
    (0 + E + E:g | subj) + (0 + E + E:g | item),
  data = d, family = Beta(),
  prior = set_prior("normal(0, .5)", class = "b") +
    set_prior("normal(1, .5)", class = "b", coef = "E") +
    set_prior("normal(0, .5)", class = "sd")
)
```

```{r m3b-summary}
fixef(m3b)
```

## M4: 2PL on the evidence weight

$\alpha_{ij} = a_j + \lambda_j \theta_i$ with $\theta_i \sim N(0, 1)$ and $\lambda_j = \exp(\log\lambda_j) > 0$ (as in Bürkner, 2021, *JSS*). Items with large $\lambda_j$ separate high and low updaters best. The person $\beta_i$ is correlated with $\theta_i$ through the shared `|p|` ID.

```{r m4}
f_2pl <- bf(
  post ~ logit_pre + (a + exp(loglam) * theta + bg * g) * E,
  a ~ 1 + (1 | i | item),
  loglam ~ 1 + (1 | i | item),
  theta ~ 0 + (1 | p | subj),
  bg ~ 1 + (1 | p | subj) + (1 | i | item),
  nl = TRUE
)

m4 <- fit_brm(
  "m4_2pl",
  control = list(adapt_delta = .95),
  f_2pl,
  data = d, family = Beta(),
  prior = set_prior("normal(1, .5)", class = "b", nlpar = "a") +
    set_prior("normal(-1.5, 1)", class = "b", nlpar = "loglam") +
    set_prior("normal(0, .5)", class = "b", nlpar = "bg") +
    set_prior("constant(1)", class = "sd", group = "subj", nlpar = "theta") +
    set_prior("normal(0, .5)", class = "sd", nlpar = "a") +
    set_prior("normal(0, .5)", class = "sd", nlpar = "loglam") +
    set_prior("normal(0, .5)", class = "sd", nlpar = "bg")
)
```

```{r m4-summary}
summary(m4)
```

```{r m4-loadings, fig.width=7, fig.height=5}
m4 %>%
  spread_draws(b_loglam_Intercept, r_item__loglam[item, ]) %>%
  mutate(lambda = exp(b_loglam_Intercept + r_item__loglam), item = as.character(item)) %>%
  left_join(item_labels, by = "item") %>%
  ggplot(aes(y = reorder(item_text, lambda), x = lambda)) +
  stat_pointinterval(.width = .95) +
  labs(x = expression("Item discrimination" ~ lambda[j]), y = NULL) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## M5: two-part (hurdle) model

Part 1 is a Bernoulli IRT model for whether the participant changed their estimate at all. (A person-level random slope on `g` in this part was not identified and kept the chains from converging, so it is omitted.) Part 2 is the M3 Beta model, fit only to trials where they did. Person effects are correlated across parts (`|p|`), as are item effects (`|i|`).

```{r m5}
f_update <- bf(upd ~ 1 + g + (1 | p | subj) + (1 | i | item), family = bernoulli())
f_size <- bf(
  post | subset(upd_l) ~ 0 + E + E:g + offset(logit_pre) +
    (0 + E + E:g | p | subj) + (0 + E + E:g | i | item),
  family = Beta()
)

m5 <- fit_brm(
  "m5_hurdle",
  f_update + f_size + set_rescor(FALSE),
  data = d,
  prior = set_prior("normal(0, 1.5)", class = "Intercept", resp = "upd") +
    set_prior("normal(0, 1)", class = "b", resp = "upd") +
    set_prior("normal(0, 1)", class = "sd", resp = "upd") +
    set_prior("normal(1, .5)", class = "b", coef = "E", resp = "post") +
    set_prior("normal(0, .5)", class = "b", coef = "E:g", resp = "post") +
    set_prior("normal(0, .5)", class = "sd", resp = "post")
)
```

```{r m5-summary}
summary(m5)
```

Person-level correlations across the two parts:

```{r m5-cor}
VarCorr(m5)$subj$cor[, "Estimate", ] %>% round(2)
```

## M6: free weights on the given and believed base rates

$c_1\,\text{logit}(b) - c_2\,\text{logit}(\tilde b) = c_1 E + (c_1 - c_2)\,\text{logit}(\tilde b)$, so the coefficient on `logit_base` is the asymmetry $c_1 - c_2$. The paper's model fixes it at 0.

```{r m6}
m6 <- fit_brm(
  "m6_free_weights",
  post ~ 0 + E + logit_base + offset(logit_pre) +
    (0 + E + logit_base | subj) + (0 + E | item),
  data = d, family = Beta(),
  prior = set_prior("normal(1, .5)", class = "b", coef = "E") +
    set_prior("normal(0, .5)", class = "b", coef = "logit_base") +
    set_prior("normal(0, .5)", class = "sd")
)
```

```{r m6-summary}
fixef(m6)
```

# Does pessimism survive?

```{r beta-across-models}
beta_summary <- function(model, par, label) {
  as_draws_df(model) %>%
    pull(!!par) %>%
    mean_qi() %>%
    transmute(model = label, beta = y, lower = ymin, upper = ymax)
}

bind_rows(
  beta_summary(m2, "b_E:g", "M2 person only"),
  beta_summary(m3, "b_E:g", "M3 + items"),
  beta_summary(m4, "b_bg_Intercept", "M4 2PL"),
  beta_summary(m5, "b_post_E:g", "M5 size | changed")
) %>%
  knitr::kable(digits = 3)
```

# Model comparison

LOO on the models fit to the same observations (M3b drops missing ages; M5 has a different likelihood).

```{r loo}
# add_criterion re-saves each fit to its cache file, so LOO is computed once
comparable <- list(M0 = m0, M1 = m1, M2 = m2, M3 = m3, M4 = m4, M6 = m6) %>%
  map(add_criterion, criterion = "loo")
loo_compare(map(comparable, ~ .x$criteria$loo))
```

Many observations have Pareto $k > 0.7$, so these PSIS-LOO estimates are unreliable. PSIS-LOO also asks how well a model predicts a held-out *trial* from a participant it has already seen.

## Grouped K-fold CV

Ten folds split by participant, identical for every model. Each held-out participant gets a new random effect drawn from the fitted population distribution (`sample_new_levels = "gaussian"`), and their 20 trials are scored jointly (`joint = "group"`). The ELPD is therefore summed over 186 participants and measures how well each model predicts a new person. M0 is omitted: it has no participant-level parameters, and it trails the other models by about 2000 ELPD in PSIS-LOO.

```{r kfold, eval = params$kfold}
library(future)
options(future.globals.maxSize = 4 * 1024^3)
plan(multisession, workers = 3)  # 3 folds at a time x 4 chains each

set.seed(2022)
fold_ids <- loo::kfold_split_grouped(K = 10, x = d$subj)

# cached per model; delete kfold_<model>.rds after changing that model
run_kfold <- function(model, name) {
  path <- file.path(fit_dir, paste0("kfold_", name, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  kf <- kfold(model, folds = fold_ids, group = "subj", joint = "group",
              sample_new_levels = "gaussian", cores = 4)
  saveRDS(kf, path)
  kf
}

# M0 has no participant parameters (and no grouping factor for joint scoring)
kfolds <- imap(comparable[names(comparable) != "M0"], run_kfold)
plan(sequential)
loo_compare(kfolds)
```

# Convergence diagnostics

```{r diagnostics}
list(M0 = m0, M1 = m1, M2 = m2, M3 = m3, M3b = m3b, M4 = m4, M5 = m5, M6 = m6) %>%
  imap_dfr(~ tibble(
    model = .y,
    max_rhat = max(rhat(.x), na.rm = TRUE),
    min_ess_ratio = min(neff_ratio(.x), na.rm = TRUE),
    divergent = sum(subset(nuts_params(.x), Parameter == "divergent__")$Value)
  )) %>%
  knitr::kable(digits = 3)
```

# Earlier draft model (superseded)

Kept for reference. It codes good news as `item_prob > pre`, which ignores event valence and the participant's own base-rate estimate (the Sharot et al. coding the paper argues against). It also models the level of the posterior rather than the update, with a Gaussian likelihood on a bounded response.

```{r draft-model, eval=FALSE}
df_long <- df %>%
  gather(key = "phase", value = "estimate", baserate, pre, post) %>%
  filter(phase == "post") %>%
  mutate(good_news = ifelse(item_prob > pre, 1, 0))

fit_cr_irt <- brm(
  bf(estimate ~ 1 + (1 | item_num) + (1 | subj_id) + good_news),
  data = df_long,
  family = gaussian(),
  chains = 4,
  cores = 4
)
```
